Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
start of section 3 edits

Computation of probabilities from hidden Markov models is well
understood and is implemented in standard software packages.
To~be able to use Haskell, we had to reimplement this standard
algorithm.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.

We~begin with our representations.
In~idiomatic Haskell, a query sequence is an immutable vector, and
we might represent an individual amino acid~$x$
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our representations are therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
type QuerySeq = Vector AA
\end{smallverbatim}

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the query sequence onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967hq}.

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{verbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{verbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type.

@@

\lstinputlisting[label=scorepathcons,caption=ScorePathCons]{scorepathcons.hs}

>>>>

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ \emph{[REPRESENTS WHAT EXACTLY?]}.

To~be able to use Haskell, we had to reimplement this standard
algorithm.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
We~begin with our representations.

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

@@

@@

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{smallverbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

@@

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type:
\begin{verbatim}
consPath :: ScorePathCons a
consPath x xs = x:xs

consNoPath :: ScorePathCons a
consNoPath _ _ = []
\end{verbatim}

@@
